---
title: "PA1_template.Rmd"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries required for script
library(lubridate)
library(dplyr)
library(lattice)
```

# Step Data Analysis

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

```{r steps}
# read data table from current working directory
data_in <- read.csv("activity.csv")
data_in$date <- as_date(data_in$date)
#
```

##Calculate the total number of steps taken per day
What is mean total number of steps taken per day?

```{r histo, echo=FALSE}
hist_data <- data_in %>% group_by(date) %>% 
        summarize(total_steps = sum(steps,na.rm = TRUE))
with(hist_data,hist(total_steps, main = "Daily Step Totals", 
     xlab="Steps",ylab = "# of Days"))
median_steps <- median(hist_data$total_steps)
mean_steps <- mean(hist_data$total_steps)
```

Mean and median of total number of steps taken per day.

Mean: `r mean_steps`

Median: `r median_steps`

##What is the average daily activity pattern?

```{r daily}
daily_data <- data_in %>% group_by(interval) %>% 
        summarize(avg_steps = mean(steps,na.rm = TRUE))
with(daily_data,plot(interval,avg_steps,type = "l",
                     main = "Average Steps by Interval",xlab="Interval",
                     ylab = "Steps"))
max_interval <- daily_data[which.max(daily_data$avg_steps),1]
```

Interval with maximum average daily step count is interval `r max_interval`.

##Imputing missing values

```{r imputed}
imputed_data <- data_in
incomplete <- sum(!complete.cases(data_in))
NAs <- is.na(data_in$steps)
interval_avg <- tapply(data_in$steps, data_in$interval, mean, na.rm=TRUE,
                       simplify=TRUE)
imputed_data$steps[NAs] <- interval_avg[as.character(data_in$interval[NAs])]
hist_imputed_data <- imputed_data %>% group_by(date) %>% 
        summarize(total_steps = sum(steps))
with(hist_imputed_data,hist(total_steps, 
                            main = "Daily Step Totals - Imputed Data",
                            xlab="Steps",ylab = "# of Days"))
median_imputed_steps <- format(median(hist_imputed_data$total_steps),
                               scientific = FALSE)
mean_imputed_steps <- format(mean(hist_imputed_data$total_steps),
                             scientific = FALSE)
```
In dataset activity.csv there are `r incomplete` missing records

Mean and median of total number of steps taken per day using imputed data:

Mean: `r mean_imputed_steps`

Median: `r median_imputed_steps`

Mean and median are now the same with the data imputed.

Comparing to non-imputed, first dataset: 

Mean: `r mean_steps`

Median: `r median_steps`

Imputing missing data raised the mean and median steps taken compared to the first step.

##Are there differences in activity patterns between weekdays and weekends?

```{r weekdays}
Weekdays <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
imputed_data$dow <- ifelse(weekdays(imputed_data$date) %in% Weekdays,"Weekday","Weekend")
weekend_data <- imputed_data %>% group_by(dow,interval) %>% 
        summarize(avg_steps = mean(steps))
xyplot(avg_steps ~ interval|dow, weekend_data, layout=c(1,2),type = "l", 
       xlab = "Interval", ylab = "Average Number of Steps")
```